---
title: "MVA Project - Factor Analysis"
author: "Chun-Jung Chen & Akshay Arora"
output:
  pdf_document: default
---

```{r include=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

```{r, include=FALSE, set.seed(333)}
knitr::opts_chunk$set(echo = TRUE,cache = T, fig.height = 5, fig.width = 8,root.dir = "~/Desktop/Supply Chain Anlaytics/Multivariate Analysis/Project/student")
```

```{r, tidy=TRUE, tidy.opts=list(width.cutoff=70), include=FALSE}
d1=read.table("student-mat.csv",sep=";",header=TRUE)
colnames(d1)[31] = "FirstMath"
colnames(d1)[32] = "SecondMath"
colnames(d1)[33] = "FinalMath"
d1 = d1[,c("school","sex","age","address","famsize","Pstatus","Medu", "Fedu", "Mjob","Fjob", "internet","guardian","traveltime","famsup","romantic","famrel",
					 "freetime","FirstMath","SecondMath","FinalMath")]


d2=read.table("student-por.csv",sep=";",header=TRUE)
colnames(d2)[31] = "FirstPort"
colnames(d2)[32] = "SecondPort"
colnames(d2)[33] = "FinalPort"
d2 = d2[,c("school","sex","age","address","famsize","Pstatus","Medu", "Fedu", "Mjob",
					 "Fjob", "internet","guardian","traveltime","famsup","romantic","famrel",
					 "freetime","FirstPort","SecondPort","FinalPort")]


d3=merge(d1,d2,by=c("school","sex","age","address","famsize","Pstatus","Medu", "Fedu",
										"Mjob", "Fjob", "internet","guardian","traveltime","famsup",
										"romantic","famrel","freetime"))

#convert data type
d3$traveltime = as.factor(d3$traveltime)
v = as.numeric(d3$traveltime)
d3$traveltime = ifelse(v==1,replace(v, which(v %in% 1),
																sample(1:15, length(which(v %in% 1)), replace = TRUE)),
			 ifelse(v==2,replace(v, which(v %in% 2), sample(16:30, length(which(v %in% 2)),
			 													replace = TRUE)),
			 	ifelse(v==3,replace(v, which(v %in% 3), sample(31:60, length(which(v %in% 3)),
			 												replace = TRUE)),
			 		ifelse(v==4,replace(v, which(v %in% 4), sample(60:90, length(which(v %in% 4)),
			 											 replace = TRUE)),v))))

d3$FinalAvg =  (d3$FinalMath + d3$FinalPort)/2
d3.q1  = d3[,c("FinalMath","famsize", "Pstatus", "famrel", "famsup")]
d3.q2  = d3[,c("FirstMath","Medu", "Fedu", "Mjob", "Fjob")]


```

```{r message=TRUE, warning=FALSE, include=FALSE, paged.print=TRUE}
#setwd("~/Desktop/Supply Chain Anlaytics/Multivariate Analysis/Project/student")

#install.packages("ggplot2")
library(ggplot2)

#install.packages("ggfortify")
library(ggfortify)

#install.packages('patchwork')
library(patchwork)

## Cluster Analysis
library(tidyverse)  # data manipulation
library(cluster)    # clustering algorithms

#install.packages('factoextra')
library(factoextra) # clustering algorithms & visualization

#install.packages("psych", lib="/Library/Frameworks/R.framework/Versions/3.5/Resources/library")
library(psych)

```

<br>

# Question 1: Deos Family conditions affect students' final grade in Math?

<br>

## Factor Analysis

\vspace{12pt}

```{r include=FALSE}
d3.q1  = d3[,c("FinalMath","famsize", "Pstatus", "famrel", "famsup")]
head(d3.q1)
str(d3.q1)
# levels(d3.q1$famsize) = c(1,0)
# levels(d3.q1$Pstatus) = c(1,0)
# #levels(d3.q1$famsup) = c(0,1)

d3.q1$famsize = as.numeric(d3.q1$famsize)

d3.q1$Pstatus = as.numeric(d3.q1$Pstatus)

d3.q1$famsup = as.numeric(d3.q1$famsup)

unique(d3.q1$famsup)
```


<br>

### Eigenvalue of variables

\vspace{12pt}

```{r}
d3q1.pc = principal(d3.q1[-1], nfactors=3, rotate="varimax")
round(d3q1.pc$values, 3)

```

\vspace{12pt}

### Cross-Loading

- Ratio for Family size
	+ $\frac{0.47}{0.10} = 4.6 \to$ ignorable 

\vspace{12pt}

- Ratio for Parents Status
	+ $\frac{0.67}{0.038} = 17.53 \to$ ignorable 

\vspace{12pt}

```{r}
d3q1.pc$loadings
```

```{r include=FALSE}
(famz1 = d3q1.pc$loadings[1,1]**2)
(famz2 = d3q1.pc$loadings[1,2]**2)
famz1/famz2

(p1 = d3q1.pc$loadings[2,1]**2)
(p2 = d3q1.pc$loadings[2,2]**2)
p1/p2
```


\newpage

### Communalities

- All variables with communalities less than 0.50 woulc identify as having sufficient explanation

- According to the result, all variables are considered as having sufficient explanation

\vspace{12pt}

```{r}
# Communalities
d3q1.pc$communality
```

\vspace{12pt}

### Scree Plot

- The reasonable factor number would be 2 or 3.

\vspace{12pt}

```{r warning=FALSE}
# Factor recommendation
fa.parallel(d3.q1[-1], show.legend = T) 

```

\newpage

### Factor scatter plot

- It is hard to find the pattern through the scattor plot.

\vspace{12pt}

```{r}
fa.plot(d3q1.pc) 
```

\newpage

### Component Diagram

- According to the Factor Analysis, there are three latern factors, which could be interpreted as parents status, faily support, and family relationship level.

```{r}
# Visualize the relationship
fa.diagram(d3q1.pc) 
```

\newpage

# Question 2: Does parents' jobs and education level influence \newline students' first period of grade in Math?


<br>

## Factor Analysis

\vspace{12pt}

```{r include=FALSE}
unique(d3.q2$Medu)
levels(d3.q2$Medu) = c(1,2,3,4,5)
unique(d3.q2$Fedu)
levels(d3.q2$Fedu) = c(1,2,3,4,5)

unique(d3.q2$Mjob)
levels(d3.q2$Mjob) = c(1,2,3,4,5)
unique(d3.q2$Fjob)
levels(d3.q2$Fjob) = c(1,2,3,4,5)


d3.q2$Medu = as.numeric(d3.q2$Medu)

d3.q2$Fedu = as.numeric(d3.q2$Fedu)

d3.q2$Mjob = as.numeric(d3.q2$Mjob)

d3.q2$Fjob = as.numeric(d3.q2$Fjob)


```


<br>

### Eigenvalue of variables

\vspace{12pt}

```{r}
d3q2.pc = principal(d3.q2[-1], nfactors=3, rotate="varimax")

round(d3q2.pc$values, 3)

```

\vspace{12pt}

### Cross-Loading

- Ratio for Mother Education level
	+ $\frac{0.677}{0.153} = 4.41 \to$ ignorable 

\vspace{12pt}

- Ratio for Father Education level
	+ $\frac{0.87}{0.015} = 56.33 \to$ ignorable 

\vspace{12pt}

- Ratio for Mother Job Type
	+ $\frac{0.93}{0.011} = 81.20 \to$ ignorable 


\vspace{12pt}

```{r}
d3q2.pc$loadings
```

```{r include=FALSE}
(ME1 = d3q2.pc$loadings[3,1]**2)
(ME2 = d3q2.pc$loadings[3,2]**2)
(ME3 = d3q2.pc$loadings[3,3]**2)
ME2/ME3
```

\newpage

### Communalities


- All variables with communalities less than 0.50 woulc identify as having sufficient explanation

- According to the result, all variables are considered as having sufficient explanation



\vspace{12pt}

```{r}
# Communalities
d3q2.pc$communality
```

\vspace{12pt}

### Scree Plot

- The reasonable factor number would be 2.

\vspace{12pt}

```{r warning=FALSE}
# Factor recommendation
fa.parallel(d3.q2[-1], show.legend = T) 

```


\newpage

### Component Diagram

- According to the Factor Analysis, there are three latern factors, which could be interpreted as parents education level, Mother's Job, and Father's Job.

```{r}
# Visualize the relationship
fa.diagram(d3q2.pc) 
```

\newpage

# Question 3: Does student's learning conditions really impact \newline students' final grade math score and Portuguese scores in average?

<br>

## Factor Analysis

\vspace{12pt}

```{r include=FALSE}
d3$FinalAvg =  (d3$FinalMath + d3$FinalPort)/2

d3.q3  = d3[,c("FinalAvg","internet", "traveltime", "romantic", "freetime")]

## Changing travel time to continuous data
v = as.numeric(d3.q3$traveltime)
d3.q3$traveltime = ifelse(v==1,replace(v, which(v %in% 1), sample(1:15, length(which(v %in% 1)), replace = TRUE)),
													ifelse(v==2,replace(v, which(v %in% 2), sample(16:30, length(which(v %in% 2)), replace = TRUE)),
																 ifelse(v==3,replace(v, which(v %in% 3), sample(31:60, length(which(v %in% 3)), replace = TRUE)),
																 			 ifelse(v==4,replace(v, which(v %in% 4), sample(60:90, length(which(v %in% 4)), replace = TRUE)),v
																 			 ))))


levels(d3.q3$internet) = c(0,1)

levels(d3.q3$romantic) = c(0,1)




d3.q3$internet = as.numeric(d3.q3$internet)

d3.q3$romantic = as.numeric(d3.q3$romantic)

#install.packages("car")
library(car)

p = powerTransform(d3.q3$traveltime, family="bcPower")
# ----------- Estimated transformation parameter = 0.1641914 

d3.q3$normtraveltime = d3.q3$traveltime**p$lambda


```


<br>

### Eigenvalue of variables

\vspace{12pt}

```{r}
d3q3.pc = principal(d3.q3[,c(2,4,5,6)], nfactors=3, rotate="varimax")

round(d3q3.pc$values, 3)


```

\vspace{12pt}

### Cross-Loading


- Ratio for Internet Assess
	+ $\frac{0.291}{0.012} = 23.74 \to$ ignorable 

\vspace{12pt}

- Ratio for freetime
	+ $\frac{0.773}{0.0164} = 47.12 \to$ ignorable 


```{r include=FALSE}
(ME1 = d3q3.pc$loadings[3,1]**2)
(ME2 = d3q3.pc$loadings[3,2]**2)
(ME3 = d3q3.pc$loadings[1,3]**2)
ME1/ME2
```

\vspace{12pt}

```{r}
d3q3.pc$loadings
```


\newpage

### Communalities


- All variables with communalities less than 0.50 woulc identify as having sufficient explanation

- According to the result, all variables are considered as having sufficient explanation



\vspace{12pt}

```{r}
# Communalities
d3q3.pc$communality
```

\vspace{12pt}

### Scree Plot

- The reasonable factor number would be 2.

\vspace{12pt}

```{r warning=FALSE}
fa.parallel(d3.q3[,c(2,4,5,6)], show.legend = T) 

```


\newpage

### Component Diagram

- According to the Factor Analysis, there are three latern factors, which could be interpreted as Spare time, Romantic relationship, and Travel time.

```{r}
# Visualize the relationship
fa.diagram(d3q3.pc) 
```



